OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd
moar wordz

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented by a similar sequence in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference. _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Unstable OTU assignments make it
difficult to use _de novo_ clustering to compare taxonomic composition of
communities between studies or to use machine learning models trained with _de
novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
approach, first performing closed reference clustering, then any sequences that
cannot be assigned to reference OTUs are clustered _de novo_ to create
additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
algorithm created the highest quality OTU assignments of all clustering methods
based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. **TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as databases or _de
novo_ OTUs and compared the performance to existing tools.

@@

**TODO: make a plan for the figures**

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering. OptiFit takes as input a list of
reference OTUs and their sequences, a list of query sequences to assign to the
reference OTUs, the sequence pairs that are within the distance threshold (e.g.
0.03), and the metric to assess clustering quality (default: MCC). Query
sequences are randomly seeded in reference OTUs, then for each sequence the
algorithm calculates the quality metric based on whether the sequences stays in
its current OTU or moves to each of the other OTUs. This process is repeated
until the quality metric stabilizes, changing by no more than 0.0001 by default,
or until a maximum number of iterations is reached (default: 100). In
closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: what determines whether a seq can't be assigned?**), and the
results only contain OTUs that exist in the original reference. In
open-reference mode, un-assigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence distance threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). As a comparison to existing software, vsearch was
used to cluster OTUs _de novo_ or with reference-based clustering to the
greengenes database. In open reference mode, OTU quality was similar between
fitting the datasets to reference OTUs with OptiFit and clustering the datasets
_de novo_ with OptiClust (**TODO: diff. in medians**). This held true for all
datasets and reference databases. OptiFit produced higher quality OTUs than
vsearch when open reference clustering against the greengenes database, with
median MCC scores of **A** and **B%** (respectively).
**TODO: open ref runtime of vsearch vs optifit vs opticlust**.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. vsearch was able to map more query
sequences to the greengenes reference than OptiFit in closed reference mode. In
terms of run time, closed reference OptiFit outperformed OptiClust by **X%**,
while OptiClust outperformed open reference OptiFit by **Y%**. **TODO vsearch
closed ref run time**.

For all datasets and clustering methods (_de novo_, open reference, and closed
reference), mothur's clustering algorithms produced higher quality OTUs than
vsearch.  In terms of runtime, OptiFit generally performed faster than vsearch
when reference clustering, while vsearch _de novo_ clustering outperformed
OptiClust.

@@

Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 80% of the sequences were created, with the
remaining sequences used for the query. Reference sequences were clustered _de
novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
OptiFit.

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
100 different random seeds. In terms of runtime, closed reference OptiFit
performed faster than OptiClust on whole datasets by **Z%**. In open reference
mode, OptiClust performed **X to Y%** faster than OptiFit only when the OptiFit
reference fraction was 30% or less. The split dataset strategy performed just as
well as the database strategy in open reference mode regardless of database
used, and outperformed the database strategy in closed reference mode by **W%**.

@@

- **TODO:** for these data, we don't see a compelling reason to use reference-based clustering over _de novo_. it was supposed to speed things up. the reason to do reference-based is if you like the ref OTUs -- e.g. for ML or downstream tools e.g. PiCrust?
- **TODO:** highlight difference between what we're doing and what previously was done.
  others use single ref seq to define an OTU, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open reference OptiFit and creates higher quality
OTUs than closed reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.
(**TODO: "if you don't have breadth, closed ref will suck." - Pat**)

The mothur algorithms produced higher quality OTUs than vsearch in
open reference, closed reference, or _de novo_ modes. However, vsearch was able
to fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a dissimilarity threshold for determining how to assign
sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
that all pairs of sequences in an OTU are within the dissimilarity threshold
without penalizing the MCC, while vsearch only requires sequences to be similar
to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
order to allow more sequences to fit to OTUs. Users who require closed reference
clustering may prefer to use vsearch if they wish to maximize the fraction of
sequences that can be fit at the cost of OTU quality. However, mothur's
OptiClust or OptiFit are preferred for _de novo_ or open reference clustering.

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. The
similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when using already-trained machine learning models to make predictions
on new data.

>>>>

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity of the reference database. Rare or novel
sequences in the sample will be lost if they are not represented by a similar
sequence in the database. _De novo_ methods cluster sequences based on their
distance to each other, without the use of an external reference. _De novo_
clustering overcomes the limitations of reference databases by considering only
sequences in the dataset, but is more computationally intensive and generates
different OTU assignments when new sequences are introduced. Unstable OTU
assignments make it difficult to use _de novo_ clustering to compare taxonomic
composition of communities between studies or to use machine learning models
trained with _de novo_ OTUs to make predictions on new data. Open reference
methods take a hybrid approach, first performing closed reference clustering,
then any sequences that cannot be assigned to reference OTUs are clustered _de
novo_ to create additional OTUs. Previous studies found that the OptiClust _de
novo_ clustering algorithm created the highest quality OTU assignments of all
clustering methods based on the Matthews correlation coefficient (MCC)
[@westcott_opticlust_2017]. As a result, we have recommended OptiClust as the
preferred method for OTU clustering whenever OTU stability is not required.

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. **TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.

@@

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering. OptiFit takes as input a list of
reference OTUs and their sequences, a list of query sequences to assign to the
reference OTUs, the sequence pairs that are within the distance threshold (e.g.
0.03), and the metric to assess clustering quality (default: MCC). Query
sequences are randomly seeded in reference OTUs, then for each sequence the
algorithm calculates the quality metric based on whether the sequence stays in
its current OTU or moves to each of the other OTUs. This process is repeated
until the quality metric stabilizes, changing by no more than 0.0001 by default,
or until a maximum number of iterations is reached (default: 100). In
closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference. In
open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, cluster the reference sequences _de novo_, then fit the query
sequences to the reference OTUs. For each dataset repeated with 100 random
seeds, we generated OTUs with OptiFit using both strategies. To compare to
existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
and with vsearch in reference-based mode against the greengenes database. All
clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the Matthews Correlation Coefficient (MCC) as described
previously [@westcott_opticlust_2017]. We calculated the fraction of query
sequences that were fit to existing OTUs in closed reference mode as an
additional measure of quality for this mode. _De novo_ clustering with OptiFit
produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
than vsearch.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), or the Ribosomal
Database Project (RDP; v16). **TODO: cite database papers?** In open reference
mode, fitting the datasets to reference OTUs with OptiFit produced OTUs of
similar quality as clustering the datasets _de novo_ with OptiClust (**TODO:
diff. in medians**). This held true for all datasets and reference databases.
OptiFit produced higher quality OTUs than vsearch when open reference clustering
against the greengenes database, with median MCC scores of **A** and **B%**
(respectively). However, open reference vsearch and OptiClust respectively ran
**X%** and **Y%** faster than OptiFit in open reference mode.

In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
in closed reference mode across any dataset/database combination. OptiFit was
able to fit **X** more query sequences to reference OTUs created with the
Greengenes and Silva databases than with RDP. vsearch was able to fit **W%**
more query sequences to the greengenes reference than OptiFit in closed
reference mode. In terms of run time, closed reference OptiFit outperformed
OptiClust by **A%** and vsearch by **B%**.

@@

A split dataset strategy was employed to assess how well OptiFit performs for
tasks where new sequences are added to existing OTUs, such as when comparing OTUs
across studies or making predictions with machine learning models.
Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 90% of the sequences were created, with the
remaining sequences used for the query. Reference sequences were clustered _de
novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
OptiFit.

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
100 different random seeds (**TODO: variation or stdev?**). In terms of runtime,
closed reference OptiFit performed faster than OptiClust on whole datasets by
**Z%**. In open reference mode, OptiClust performed **X to Y%** faster than
OptiFit only when the OptiFit reference fraction was 30% or less. The split
dataset strategy performed just as well as the database strategy in open
reference mode regardless of database used, and outperformed the database
strategy in closed reference mode by **W%**.

@@

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP, which has only about 16,000 V4
16S sequences compared to about 170,000 in Silva and greengenes. We recommend
users who require an independent reference database opt for large databases with
good coverage of microbial diversity. Since OptiClust performs faster than open
reference OptiFit and creates higher quality OTUs than closed reference OptiFit
with the database strategy, we recommend using OptiClust rather than fitting to
a database whenever stable OTUs are not required for the study at hand.

The mothur algorithms produced higher quality OTUs than vsearch in open
reference, closed reference, or _de novo_ modes. However, vsearch was able to
fit more sequence into OTUs than OptiFit in closed reference mode. While both
mothur and vsearch use a dissimilarity threshold for determining how to assign
sequences into OTUs, vsearch is more permissive than mothur. Mothur's OptiFit
and OptiClust use all of the sequences to define an OTU, requiring that all
pairs of sequences (including reference and query sequences) in an OTU are
within the distance threshold without penalizing the MCC. In contrast,
vsearch only requires each query sequence to be similar to the single sequence
that seeded the OTU. In this way, vsearch sacrifices OTU quality in order to
allow more sequences to fit to OTUs. Users who require closed reference
clustering to the greengenes database may prefer to use vsearch if they wish to
maximize the fraction of sequences that can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. The
similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when using already-trained machine learning models to make predictions
on new data or comparing OTUs across studies. However, when stable OTUs are not
required, we recommend using OptiClust for _de novo_ clustering over the split
strategy with OptiFit since OptiClust simpler to execute but performs similarly.
